Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-07-17 response-to-review-new.tex

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Response to Referee’s report of SIM submission 13-0382.R1 entitled "Comparing and
combing biomarkers as principle surrogates for time-to-event clinical
endpoints"}
\author{Erin E. Gabriel, Michael C. Sachs, Peter B. Gilbert}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
We are happy to have the opportunity to revise our manuscript. We have put forth considerable effort to address reviewer 2's comments and we feel that the paper has improved greatly.

We have made some critical revisions to the methods section, and have modified the real-data example to better illustrate our proposed methods.  We have developed a new summary statistic that does not require the selection of a specific time. During the development of this new summary statistic, we noticed a convergence issue for the multivariate PS estimands, that can lead to poor efficiency in practice. This issue was previously masked by our truncation of the STG for a given time point. We now consider using a multivariate BIP, one independent BIP for each candidate PS, that appears to solve the convergence issues.

We have spent several months attempting to locate an adequate new example data set. Although we were unable to do so, we tried several, including the Mr. FIT data, and several unpublished small immune marker trial of NIAID. In each case there was inadequate baseline information to support the estimation of the counterfactual estimands. However, in running each of these examples, we determined that the use of truncated normal in the simulations did not fit what we were suggesting in practice for the example. We have modified these simulations to use censored normal generation and censored imputation. Below we outline the specific changes.

\section{Comments from Reviewer 2}
\begin{quotation}
I have to admit that I am somewhat disappointed with this revision.

Only minor changes in the essential section 2 have been made.

The paper is still not sufficiently self-contained and one has to read at least [1] and [10] to have a chance at understanding what’s going on in the paper. Both [1] and [10] are very nice and interesting papers by the way.

The extension from [10] in terms of modelling is straightforward and minor.

The tools developed to assess and select one or more surrogates are problematic at least to me as they include the subjective choice of one or more time points at which performance is evaluated. Also, the actual selection of an optimal subset of surrogates is not well described.

It would be ideal with an example where these extensions really mattered.
\end{quotation}

\section{Response to Reviewer 2}
\begin{enumerate}
\item We reorganized the methods into sections 2 and 3. Section 2 now contains the notation and background information on the concept of principal surrogacy. This section provides a more thorough introduction to the methods so that uninitiated readers may better understand without referring to [1] or [10]. 
\item The remaining subsections of the methods have been reorganized so that the concepts flow more naturally from defining estimands of interest, to assumptions, to modeling, to estimation, to summary statistics, and finally to PS comparison. 
\item We have added an additional subsection where we introduce the integrated standardized total gain to address reviewer 2's concerns about the subjective choice of time points.  The integrated STG does not require the analyst to specify a time point, but rather averages the time dependent STG with respect to the distribution of event times. We suggest estimating the marginal distribution of event times with the Kaplan-Meier estimator. We evaluate the integrated STG in the simulations and demonstrate its operating characteristics. The integrated STG tends to have slightly higher power for comparison of candidate PS as compared to the STG at a single time point. 
\item In regard to the reviewer's comment about selecting the optimal subset of surrogates, we have revised section 3.7 to more clearly describe the hypothesis tests that are used to rule out useless PS from consideration and for comparing and ranking candidates with some value. 
\item We have modified the example to better illustrate our extensions. We created a third candidate PS from a linear combination of the two real candidate PS such that the PS has some value as a surrogate, but with a significantly smaller STG. In this case the test for the difference in STG yields evidence of a difference between the two surrogates where standard hypothesis testing could not. This makes the example more like a simulation, in that we modified the data to generate the desired results, but we feel that it does provide a better illustration of where our proposed methods actually matter. 
\item We have added to the multivariate simulations the investigation of the use of a multivariate BIP. 
\item We have added updated the simulations to reflect our suggested censoring of a candidate surrogate, as done in the example, rather than truncation. 
\end{enumerate}

We thank you again for the thoughtful review; the paper has improved greatly thanks to your efforts.

\vspace{2em}

\hfill{Sincerely,}
\vspace{2em}

\hfill{Erin E. Gabriel}

\end{document}

### 2014-07-17 supp-power-table.tex
Revisions, vector W, still needs new simulation tables

\documentclass[times, doublespace]{simauth}

\usepackage{moreverb}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 \usepackage[numbers, square, sort&compress]{natbib}

\begin{document}
% The following line would print the thesis in 
%opening

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Kidney Research Institute, University of Washington\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

\section{Introduction} 
A valid surrogate endpoint can be used as a primary outcome for evaluating treatments in phase I-II trials, by virtue of providing reliable prediction of treatment effects on a clinical endpoint of interest. In 2002, \citet{Frangakis02} introduced the principal stratification framework and a definition of a principal surrogate (PS). Several methods for evaluation of principal surrogates have been developed under this framework, eg. \citep{Taylor05, Follmann06, Gilbert08, Wolfson10, Li10, Huang11, Huang12, Zigler12, Gabriel13}.  We extend this literature in several ways.

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero. A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement. Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect. This association can be displayed by the marginal-CEP versus candidate PS curve \citep{Gilbert08}. Evaluation of multivariate principal surrogates is complicated by the increased dimension.

The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP curve difficult to display and interpret for multivariate candidate PS. If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function.

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS . Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standardized total gain \citep{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial. The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring. Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.

In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.  In Section 3.1, we describe the time-dependent risk estimands and outline some assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7, we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combinations of the two candidates are not improvements over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

\section{Notation and Criterion for Principal Surrogacy} \label{DEF}
Let $Z$ be the treatment indicator, 0 for control/non-active treatment and 1 for treatment. Let $W$ be a vector of baseline measurements taken prior to randomization. In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual. Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$. Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$. Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$. Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$. Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$. Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential outcomes for subject $i$.

All $S_j$ are measured at the same fixed post-randomization time point $\tau$. If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition. For this reason, subjects with $T< \tau$ are excluded from the evaluation cohort. Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$. We assume the full set of observed and counterfactual outcomes are independently and identically distributed over the trial subjects $i= 1, \ldots n$. Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively. Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$. Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should strongly modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, the wide variability criterion. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$. ACN is a sensible condition that can help eliminate candidate PS, but is not a sufficient condition, as was shown in \citet{Vanderweele11}. The combination of wide effect modification and ACN are often used for evaluation in the current literature. The wide effect modification criterion may also suggest a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.

In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate. marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$. The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.

For a univariate candidate PS ACN can be displayed using the joint CEP, or marginal CEP under case CB.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. As well, the definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, ACN may not guarantee this is the case \citep{Vanderweele11}. We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional on a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}. Like ACN, the classification accuracy measures allow for the quantification of the agreement between the treatment effect on the clinical outcome and the predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ACN, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves versus the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary but not sufficient for a set of biomarkers to have value as a PS. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

 For a multivariate PS the concept of wide variation in the CEP over subgroups defined by the set of biomarkers is clear. However, this is again difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of the wide variation criterion. We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate wide variation. We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and wide variation in a single measurement. Details for standardized total gain are given in Section \ref{STG}.  As well, if one assumes a parametric model for risk conditional on a set of biomarkers then a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model. The specifics of this test for our assumed model are given in Section \ref{CRE}.

After a set of candidate multivariate or univariate candidate surrogates is determine to have some value as PS, it is of interest to rank them and test to determine if any one candidate is significantly better than others. ANC can only classify candidates into two groups. Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates. We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference. These are outlined in Sections \ref{STG} and \ref{ISTG}.

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can be defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal clinical risk up to time $\tau$: $T(1) \leq\tau$ if and only if $T(0) \leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
\end{itemize} 
Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z\in \{0,1\}$.

Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model. We will assume a Weibull model, the probability density function (pdf) for which we will denote as $g(\cdot|\cdot,\cdot)$. We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model. The density is given by:
\begin{eqnarray*}
g(t|\gamma, \beta, z, s_{1,1}, \ldots,s_{1,J}, y)=\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})^{y} Q_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J}),
\end{eqnarray*}
where $Q_z(t|\gamma, \beta, s_{1,j})$ is the conditional Weibull survivor function for treatment arm $Z=z$ and $\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})$ is the conditional Weibull hazard function. We assume the risk estimands are of the form:
\begin{itemize}
\item A5: $risk_{z}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) \equiv 1-\exp(-(t/\exp(\gamma_{z0}+\sum_j{\gamma_{zj}s_{1,j})})^{(\exp(\beta_{0}))}$
\end{itemize}
for $z\in \{0,1\}$. We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials. In order for both assumptions A3 and A5 to hold simultaneously, we can only consider censoring or event times greater than $\tau$. This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.

One can test the null hypothesis of no surrogate value or no variation in the marginal-CEP, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{j}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.

\subsection{Trial Augmentation and Estimation} \label{EST}
The assumed parametric form of the risk estimands, A5, condition on the $S_{j}(1)$, which are missing for all subjects assigned non-active treatment $z=0$. Several papers have suggested the use of baseline covariates to account for the missing $S_{j}(1)$ values, a strategy called the ``baseline immunogenicity predictor'' (BIP) augmented trial design \citep{Follmann06, Gilbert08}. A useful BIP, $W$, is highly correlated with the $S_{j}(1)$ which allows for good prediction of the missing $S_i(1)$. Although other trial augmentations have been proposed in the PS literature, the only augmentation available in our motivating data set is BIP, so it will be our focus. Using the BIP trial augmentation the observed likelihood given Assumption A5 is:
\begin{eqnarray*}
&&\mbox{L}(\beta,\gamma,\nu) \equiv \prod^{n}_i\left\{g(T_i|\gamma, \beta, Z_i, s_{1,1,i},\ldots,s_{1,J,i}, Y_i)\right\}^{\prod_{j=1}^{J}{M_{j,i}}}\\ 
&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, s_{1,l+1},\ldots, s_{1,J}, Y_i)dF_{S_{l+1}(1),\ldots,S_{J}(1)|W}(s_{1,l+1},\ldots,s_{1,J})\right\}^{\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\}}\\ 
&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1}, \ldots, s_{1,J}, Y_i)dF_{S_{1}(1),\ldots, S_{J}(1)|W}(s_{1,1},\ldots, s_{1,J})\right\}^{\{\prod_{j=1}^{J}{(1-M_{j,i})}\}}.
\end{eqnarray*}

This likelihood could be solved by direct optimization if we assumed a parametric model for $S_j(1)|W$. However, as proposed in \citet{Pepe91}, consistent estimation of outcome model parameters can be achieved using an estimated distribution of $S_j(1)|W$; this is called estimated maximum likelihood. Following \citet{Huang11}, we assume the semi-parametric location-scale model of \citet{Heagerty99} for $F_{S_{j}(1)|W}$. Considering first a univariate candidate surrogate, the semi-parametric model is defined as $F_{S_{j}(1)|W} \sim F_j[\{s_{1,j} -\mu_j(w)\} / \sigma_j(w)]= F_j(\varsigma_j)$, where $F_j$ is the CDF of the univariate residuals $\varsigma_j$ from fitting the location and scale parameters $\mu_j(w)$ and $\sigma_j(w)$, respectively.  We define $\mu_j(w)$ and $\sigma_j(w)$ as parametric functions of the baseline variable(s) $W$, $\mu_j(w) = \gamma_j'w$ and $\log\{\sigma_j(w)\} = \eta_j'w$. We can estimate these parameters by solving the equations given in \citet{Heagerty99}:
\begin{eqnarray}
&&\sum_{k=1}^{n_V} {w_{k} (s_{(1,j,k)} - \gamma_j' w_{k} )\over \{\exp(\eta_j'w_k)\}^2} =0\\ \label{stuff}
&&\sum_{k=1}^{n_V} {w_{k}\{(s_{(1,j,k)} - \gamma_j' w_{k})^2 - \{\exp(\eta_j'w_k)\}^2\} \over \{\exp(\eta_j'w_k)\}^2}=0,\label{stuff2}
\end{eqnarray}
among those treated subjects with all $S_{(j,k)}(1)$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.

Fitting the set of parameter equations for the validation subjects yields $n_V$ residuals, $\varsigma_{j,k} \equiv S_{j,k}(1) - \hat{\gamma_j}'w_{k} + \exp(\hat{\eta_j}'w_{k})$. Missing counterfactual $S_{j}(1)$ values can be imputed using the parameter estimates and the $n_V$ residuals by, $S_{j,i,k}^*(1) = \hat{\gamma_j}'w_{i} + \exp(\hat{\eta_j}'w_{i}) \varsigma_{j,k}$ for $n_V$ total imputations for each missing $S_{j}(1)$ value. We can then use these imputed values $S_{j,i,k}^*(1)$ to estimate the EML contribution for subject $i$ by $\int g(t_i|\gamma, \beta, z_i, s_{1,1}, \ldots, s_{1,J}, y_i)dF_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$, by the empirical integral ${\left(1/n_{V}\right)}\sum_k^{n_V}{g_z(t_i|\gamma, \beta,z_i, S_{1,k}^*(1), \ldots,S_{J,k}^*(1),y_i)}$. Missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be imputed in the same way. In cases where the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the biased sampling.

To accommodate multivariate surrogates, the same location-scale model is fit for each of the univariate candidate PS separately in the set of validation subjects. Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model, empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation. This is an important limitation of this method. The BIP need not be based on a single baseline measurement or have the same correlation with each candidate, but the BIP must be the same in each location-scale model for the estimated likelihood to be proper using our procedure. This is not a requirement to compare univariate candidate surrogates not included in the same model. This gives us an estimated log likelihood of:
\begin{eqnarray*}
&&\mathit{l}(\beta,\gamma, \hat{\nu})=\sum_{i} log(g_z(T_i|\gamma, \beta,Z_i, s_{1,1,i}, \ldots,s_{1,j,i},Y_i))*\left\{\prod_{j=1}^{J}{M
_{j,i}}\right\}+\\
&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, S^{*}_{l+1,i,k}(1),\ldots, S^{*}_{J,i,k}(1), Y_i))}\right\}*\\
&&\left\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\right\}+\\
&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g_z(T_i|\gamma, \beta,Z_i, S_{1,i,k}^*(1), \ldots,S_{j,i,k}^*(1),Y_i))}\right\}*\left\{\prod_{j=1}^{J}{(1-M
_{j,i})}\right\}\cdot
\end{eqnarray*}

The asymptotic distributional results of \citet{Pepe91} for general EML estimators do not carry over to our setting, due to the zero probability of observing $S_{j}(1)$ in placebo recipients. We suggest the bootstrap for inference. The assumed form of the risk estimates, A5, can be estimated using the parameters from solving the EML and the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, can be estimated directly from the risk estimates.

>>>>

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $STG(t)$ tests for $t=3$ years \label{Power}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.20$& $STG(3)^{*}=0.28$  & $STG(3)^{*}=0.31$\\
  \hline
$STG(3)=0.1$ &0.8 & 5.5 & 17.0 & 16.8 & 28.2 & 71.0 \\ 
   &0.5 &5.8 & 11.9 & 14.2 & 45.6 & 73.7 \\ 
$STG(3)=0.24$ &0.8&  & 5.1 & 7.9 & 7.5 & 13.7 \\ 
   &0.5 &  & 5.5 & 8.9 & 7.4 & 13.3 \\ 
$STG(3)=0.20$&0.8 &  &  & 5.3 & 52.5 & 96.4 \\ 
   &0.5&  &  & 6.3 & 45.5 & 93.9 \\ 
$STG(3)^{*}=0.28$&(0.8, 0.8) &  &  &  & 5.4 & 7.7 \\ 
   &(0.5, 0.5) &  &  &  & 5.7 & 8.9 \\ 
$STG(3)^{*}=0.31$&(0.8, 0.8)&   &  &  &  & 5.1 \\ 
   &(0.5, 0.5)&  &  &  &  & 4.2 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $STG_1(3)-STG_2(3)=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
\end{tabular}
\end{center}
\end{sidewaystable}
